---
title: "The **seqtsne** package"
author: "Joseph R Boyd"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteIndexEntry{Overview and Use Cases}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[UTF-8]{inputenc}
---

<style type = "text/css"> em { font-weight: bold; } </style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    dpi=60
)
```

# Synopsis
seq - tee - snee

`r Githubpkg("jrboyd/seqtsne")` 

This package is in active development.  *Expect changes in function and
parameter names!*

# Features

* tsne on ChIP-seq data

# Functions

*Eventually*

* `sts*` - most primary user facing functions are prefixed by sts



# Installation and Loading

## From github

```{r bioc install, eval=FALSE}
devtools::install_github("jrboyd/seqtsne")
```

## Load the library
```{r load seqtsne, message=FALSE}
library(seqtsne)
```

## Load optional useful libraries
```{r load optional libs, message = FALSE}
theme_set(theme_classic())
```

# Running t-sne

## Set parameters

These parameters will determine how multiple functions behave

```{r parameters}
n_points = 6
color_mapping = c("H3K4me3" = "forestgreen",
                  "H3K27me3" = "firebrick1")
options("mc.cores" = 2)
perplexity = 15
```

The two critical inputs are:

1. data.table containing bigWig files and
cell and mark metadata.
2. GRanges of regions to retrieve data for and perform t-sne on.

```{r setup files and configure}
# input_dirs = c(
#     dir("/slipstream/galaxy/uploads/working/qc_framework/output/", 
#         pattern = "K4.+_pooled$", full.names = TRUE),
#     dir("/slipstream/galaxy/uploads/working/qc_framework/output/", 
#         pattern = "K27.+_pooled$", full.names = TRUE)
# )
# #assemble peak calls and sample
# peaks = dir(input_dirs, pattern = "pooled_peaks.narrowPeak", full.names = TRUE)
# olaps = ssvOverlapIntervalSets(easyLoad_narrowPeak(peaks), ext = 100)
# olaps = olaps[rowSums(as.data.frame(mcols(olaps))) > 3] 
# length(olaps)
# set.seed(0)
# # query_gr = sample(olaps, 100)
# query_gr = olaps
# query_gr = resize(query_gr, 5000, fix = "center")
# query_gr$id = names(query_gr)
# #create config from bws
# bws = dir(input_dirs, pattern = "pooled_FE.bw", full.names = TRUE)
data("query_gr")
bw_files = dir(system.file('extdata', package = "seqtsne"), 
               full.names = TRUE, pattern = ".bw$")
cfg_dt = data.table(file = bw_files)
cfg_dt[, c("cell", "mark") := tstrsplit(basename(file), "_", keep = 1:2)]
```

t-sne requires a wide-matrix view where each row is a point (observation) and 
each column is a dimension (attribute) of the data.

```{r prepare input}
tsne_input = stsFetchTsneInput(cfg_dt, query_gr)
```

There's a bit of art to running t-sne, see this 
[hands-on explanation](https://distill.pub/2016/misread-tsne/) to get oriented.

Be sure to set `mc.cores` using `options()` or provide the `n_cores` 
parameter to speed up processing time.

```{r run tsne}
tsne_res = stsRunTsne(tsne_input$bw_dt, perplexity = perplexity)
```

# Plotting

## Describing the space

A basic plot facetted by the 3 cell lines.

```{r basic plot, fig.width=13, fig.height=4}
ggplot(tsne_res, aes(x = tx, y = ty, color = cell)) + 
    geom_point() +
    facet_wrap("cell")

```

Interpreting the t-sne landscape can be difficult but it is critical for
effectiveness of the visualization.  We can plot representative profiles in
the t-sne landscape.  

The resolution of these images is controlled by `n_points`.  Use `xrng` and
`yrng` to specify a region to view in detail.  A strength of t-sne is that
it maps both broad trends and finer relationships effectively.

```{r, fig.width=5, fig.height=5}
img = prep_images(tsne_input$bw_dt, tsne_res, x_points = n_points,
                    line_color_mapping = color_mapping)
img_res = plot_summary_raster(img$image_dt, x_points = n_points, min_size = 0)
img_res
```

The same idea as before but now facetted by cell.

```{r, fig.width=7.5, fig.height=3}
img_byCell = prep_images(tsne_input$bw_dt, 
                         tsne_res, 
                         x_points = n_points,
                         line_color_mapping = color_mapping, 
                         facet_by  = "cell")
img_byCell_res = plot_summary_raster_byCell(img_byCell$image_dt, x_points = n_points)
img_byCell_res
```

## Velocity

Arrows identity how regions behave between two groups.  Each arrow is drawn 
from the postion in `cell_a` to position in `cell_b` with color mapped to 
angle to aid in pattern detection.

```{r, fig.width=8, fig.height=4}
vel_dt = prep_velocity(tsne_res, cell_a = "HUES48", cell_b = "HUES64")
vel_res = plot_velocity_arrows(vel_dt)
vel_res
```

Individual arrows can be overwhelming, we can also employ a binning strategy
to summarize the average destination in t-sne space for groups of points.
Each arrow here is drawn to the average desination (`cell_b`) for all nearby 
points (`cell_a`).

```{r, fig.width=4, fig.height=3}
plot_regional_velocity(tsne_dt, cell_a = "HUES48", cell_b = "HUES64", x_points = n_points)
```

We can also layer the binned arrows on top of the binned profiles.

```{r, fig.width=4, fig.height=3}
plot_regional_velocity(tsne_dt, cell_a = "HUES48", cell_b = "HUES64", 
                       x_points = n_points, p = img_res)#, p = img_res$plot)
```

## Other

Individual profiles can also be viewed.

```{r, fig.width=6, fig.height=3}
plot_profiles_selected(tsne_input$bw_dt, 
                       qcells = c("HUES48", "HUES64"), 
                       id_to_plot = query_gr$id[1:5], 
                       color_mapping = color_mapping)
```


