---
title: "The **seqtsne** package"
author: "Joseph R Boyd"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
    %\VignetteIndexEntry{Overview and Use Cases}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

<style type = "text/css"> em { font-weight: bold; } </style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    dpi=60
)
```

# Synopsis
seq - tee - snee

`r Githubpkg("jrboyd/seqtsne")` 

This package is in active development.  *Expect changes in function and
parameter names!*

# Features

* tsne on ChIP-seq data

# Functions

*Eventually*

* `sts*` - most primary user facing functions are prefixed by sts



# Installation and Loading

## From github

```{r bioc install, eval=FALSE}
## try http:// if https:// URLs are not supported
# if (!requireNamespace("BiocManager", quietly=TRUE))
#     install.packages("BiocManager")
# BiocManager::install("seqtsne")
devtools::install_github("jrboyd/seqtsne")
```

## Load the library
```{r load seqtsne, message=FALSE}
library(seqtsne)
```

## Load optional useful libraries
```{r load optional libs, message = FALSE}
library(GenomicRanges)
library(data.table)
library(cowplot)
library(seqsetvis)
theme_set(cowplot::theme_cowplot())
```

# Running t-sne

## Set parameters

These parameters will determine how multiple functions behave

```{r parameters}
n_points = 6
color_mapping = c("H3K4ME3" = "forestgreen",
                  "H3K27ME3" = "firebrick1", 
                  "H3K4AC" = "blue", 
                  "H3K27AC" = "purple")


options("mc.cores" = 20)
perplexity = 500
```

The two critical inputs are:

1. data.table containing bigWig files and
cell and mark metadata.
2. GRanges of regions to retrieve data for and perform t-sne on.

```{r setup files and configure}
input_dirs = c(
    dir("/slipstream/galaxy/uploads/working/qc_framework/output/", 
        pattern = "K4.+_pooled$", full.names = TRUE),
    dir("/slipstream/galaxy/uploads/working/qc_framework/output/", 
        pattern = "K27.+_pooled$", full.names = TRUE)
)
#assemble peak calls and sample
peaks = dir(input_dirs, pattern = "pooled_peaks.narrowPeak", full.names = TRUE)
olaps = ssvOverlapIntervalSets(easyLoad_narrowPeak(peaks), ext = 100)
olaps = olaps[rowSums(as.data.frame(mcols(olaps))) > 3] 
length(olaps)
set.seed(0)
# query_gr = sample(olaps, 100)
query_gr = olaps
query_gr = resize(query_gr, 5000, fix = "center")
query_gr$id = names(query_gr)
#create config from bws
bws = dir(input_dirs, pattern = "pooled_FE.bw", full.names = TRUE)
cfg_dt = data.table(file = bws)
cfg_dt[, c("cell", "mark") := tstrsplit(basename(file), "_", keep = 1:2)]
```

```{r make_wide}
cfg_dt[, mark := paste(cell, mark, sep = "_")]
cfg_dt$cell = "combined"
color_mapping = rep(color_mapping, 3)
names(color_mapping) = paste(rep(c("MCF10A", "MCF7", "MDA231"), each = 4), names(color_mapping), sep = "_")

group_mapping = sub("_.+", "", names(color_mapping))
names(group_mapping) = names(color_mapping)

```

t-sne requires a wide-matrix view where each row is a point (observation) and 
each column is a dimension (attribute) of the data.

```{r prepare input}
tsne_input = fetch_tsne_mat(cfg_dt, query_gr)
```

There's a bit of art to running t-sne, see this 
[hands-on explanation](https://distill.pub/2016/misread-tsne/) to get oriented.

Be sure to set `mc.cores` using `options()` or provide the `n_cores` 
parameter to speed up processing time.

```{r run tsne}
tsne_res = run_tsne(tsne_input$tsne_mat, perplexity = perplexity)
```

# Plotting

## Describing the space

A basic plot facetted by the 3 cell lines.

```{r basic plot, fig.width=13, fig.height=4}
ggplot(tsne_res, aes(x = tx, y = ty, color = cell)) + 
    geom_point() +
    facet_wrap("cell")

```

Interpreting the t-sne landscape can be difficult but it is critical for
effectiveness of the visualization.  We can plot representative profiles in
the t-sne landscape.  

The resolution of these images is controlled by `n_points`.  Use `xrng` and
`yrng` to specify a region to view in detail.  A strength of t-sne is that
it maps both broad trends and finer relationships effectively.

```{r, fig.width=5, fig.height=5}
img = make_tsne_img(tsne_input$bw_dt, tsne_res, n_points = n_points, 
                    line_colors = color_mapping, 
                    group_mapping = group_mapping, 
                    force_rewrite = F)
img_res = plot_tsne_img(img$images_dt, n_points = n_points, min_size = 0)
img_res$plot
```

The same idea as before but now facetted by cell.

This makes no sense because we've combined all the cells!

```{r, fig.width=5, fig.height=5}
img_byCell = make_tsne_img(tsne_input$bw_dt, tsne_res, n_points = n_points, 
                    line_colors = color_mapping, group_mapping = group_mapping, 
                    facet_by = "cell", force_rewrite = F)
img_byCell_res = plot_tsne_img_byCell(img_byCell$images_dt, img_byCell$tsne_dt, n_points = n_points)
img_byCell_res$plot

```

## Velocity

Velocity is not applicable when combinding cell data in this way.

## Other

Individual profiles can also be viewed.

```{r, fig.width=6, fig.height=3}
indi_dt = copy(tsne_input$bw_dt)
indi_dt[, c("cell", "mark") := tstrsplit(mark, "_")]
indi_col = color_mapping[1:4]
names(indi_col) = sub(".+_", "", names(indi_col))
plot_profiles_selected(indi_dt, 
                       qgr = query_gr[1], 
                       qcells = c("MCF10A", "MCF7", "MDA231"), 
                       tss_ids = names(query_gr)[1], 
                       color_mapping = indi_col)
```


